| Environmental and Development Economics\n| Module 2 - How does development affect the environment?

### 2024-11-11 lectures/module1/module1.Rmd
added .Rmd files

- **Class Time/Location:** Tues/Thurs 11:45am-1:25pm, Ruttan 119
\vfill
- **Office Hours:** Thursdays 1:30pm-2:30pm, Ruttan 337D
\vfill
- **Course website:** https://github.com/rmadhok/enviro-dev-grad
  - lectures, assignments, syllabus
  - Try to skim reading(s) beforehand
\vfill
- **Assignments**: Upload through Canvas
\vfill

@@

- $n$ identical agents with utility from consumption, EQ, and health
- Initial wealth $y_{0}$, health $h_{0}$, environmental equality $e_{0}$
- health depends on self-protection, $s$, and $e$
- Assume perfect markets (i.e. no externalities)

@@

## The Social Planner

- In first best, social planner sets MB = MC
  - where $MC_{e} = \frac{\partial c_{e}}{\partial \Delta e}$ and $MC_{s} = \frac{\partial c_{s}}{\partial \Delta s}$
\vfill
- But to aggregate over $n$, we must assume:
  - No preferences of her own
  - No market failures
  - Can observe true MWTP
  - Anything else?
\vfill
- Do these hold in LMICs?
<!-- political economy and market failures -->
<!-- what about externalities? this doesn't differ between HICs and LMICs. But they may be exacerbated by or interact with market failures -->

@@

- Set the stage: 
  - how does environment affect development ($\frac{\partial h}{\partial e}$) (week 2)
  - how does development affect the environment (week 3)

@@

- Goal: where can you make a contribution?

@@

- Social planner includes own utility weights social welfare function
  - i.e. corruption
\vfill
- Many examples from LMICs
  - pollution (Duflo et al., 2013)
  - deforestation (Burgess et al., 2012; Viera et al., 2024)
  - human-wildlife conflict (Madhok et al., 2024)
\vfill
- Leads to second best policy (inefficient)
\vfill

@@

- This is partially a couse on development economics
  - About market failures: land, labor, credit, etc.
\vfill
- Implication for us: revealed $MWTP_{e} \neq$ first best $MWTP_{e}$
\vfill
- Example: weak property rights $\rightarrow$ underinvestment in $e$
  - Underestimate $MWTP_{e}$ from observed data
  - RCT evidence from crop-burning PES contracts: Jack et al. (2024)
\vfill

>>>>

- **Class Time/Location:** Tues/Thurs 11:45am-1:25pm, Ruttan 119
  - Last class: Oct 17th

\vfill
- **Office Hours:** Thursdays 1:30pm-2:30pm, Ruttan 337D
\vfill
- **Course website:** https://github.com/rmadhok/enviro-dev-grad
  - lectures, assignments, syllabus
  - Try to skim reading(s) beforehand
\vfill
- **Assignments**: Upload through Canvas
\vfill

@@

- $n$ identical agents with utility from consumption, EQ, and health
- Initial individual wealth $y_{0}$, health $h_{0}$, environmental equality $e_{0}$
- health depends on self-protection, $s$, and $e$
- Assume perfect markets (i.e. no externalities) to benchmark first-best

@@

## The Fist Best Outcome

- Agent sets ratio of marginal costs = ratio of marginal benefits
$$\frac{MWTP_{e}}{MWTP_{s}}=\frac{\frac{\partial c_{e}}{\partial \Delta e}}{\frac{\partial c_{s}}{\partial \Delta s}}$$

\vfill
- For social planner to aggregate over $n$, we must assume:
  - No preferences of her own
  - Can observe true MWTP
  - Anything else?
\vfill
- Do these hold in LMICs?
<!-- political economy and market failures -->
<!-- what about externalities? this doesn't differ between HICs and LMICs. But they may be exacerbated by or interact with market failures -->

@@

- Set the stage: 
  - how does environment affect development ($\frac{\partial h}{\partial e}$) (module 2)
  - how does development affect the environment (module 3)

@@

- Goal: where can you make a contribution?
  - E.g. Lots of work on $\frac{\partial h}{\partial e}$, but very little on $\frac{\partial \Delta y}{\partial e}$

@@

- Social planner includes own utility weights social welfare function
  - i.e. corruption
\vfill
- Many examples from LMICs
  - pollution (Duflo et al., 2013)
  - deforestation (Burgess et al., 2012)
  - human-wildlife conflict (Madhok et al., 2024)
\vfill
- Leads to second best policy (inefficient)
\vfill

@@

- This is partially a course on development economics
  - About market failures: land, labor, credit, etc.
\vfill
- Implication for us: revealed $MWTP_{e} \neq$ first best $MWTP_{e}$
\vfill
- Example: weak property rights $\rightarrow$ underinvestment in $e$
  - Underestimate $MWTP_{e}$ from observed data
  - RCT evidence from crop-burning PES contracts: Jack et al. (2024)
\vfill

### 2024-11-11 lectures/module2/module2.Rmd
added .Rmd files

---
title: "| Environmental and Development Economics\n| Week 2 - How does development affect the environment?
  \n"
author: |
  | Raahil Madhok
  | UMN Apllied Economics
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: Montpellier
  slidy_presentation: default
  ioslides_presentation: default
classoption: aspectratio=169
header-includes:
- "\\AtBeginDocument{\\title[Week 2 - development and environment]{Environmental and Development
  Economics \\\\ Week 2 - How does development affect the environment?}}"
- \input{header.tex}
---

@@

- State things that are due.

@@

- **Guiding question:** how does economic development affect the environment?
\vfill
- Descriptive overview
\vfill
- Channel I: Income effects
  - Changes in consumption
  - Energy
  - Diet
\vfill
- Channel II: Technology and Infrastructure
  - Agricultural productivity
  - Infrastructure
\vfill
- Channel III: Institutions
  - Later in the course

@@

@@

- What variation in "development" can approximate the experiment?
  - what proxies development?
  - what does the proxy miss?
  - what (non-random) variation isolates causal relationship?
  - what does that variation miss?

@@

- Poverty alleviation may $\uparrow$ demand for resource-intensive good
\vfill
- Or, it may $\uparrow$ demand for conservation

\vfill
- Empirical challenge: household income is endogenous
\vfill
- **This paper:** Exploit RDD eligibility to study impact of CCT on forests

@@

\vfill
- If we control for OVB, will panel fixed effects work? 
  - No: small year-to-year income changes do not reflect ``development''
  - Predictable short-term income fluctuations different than permanent change

@@

- **Main variables:** village lat/lon, eligibility index, forest cover
\vfill
- **Sample size:** 105,749 villages
\vfill
- **Unit of analysis:** only village centroids provided?
  - How do we estimate village-level impacts? Thiessen polygons
  - Area around a point where every location closer to the point than to all others
  - Problems?

@@

## Empirical Strategy

- OLS:
$$\Delta f_{i} = \alpha + \delta E_{i} + \beta'X_{i} + \epsilon_{i}$$
where $E_{i}=1$ if index $I_{i} > -1.2$

\vfill
- RDD: Sample window $-2 < I < -0.2$
\vfill
- Fuzzy RDD/IV:
$$\Delta f_{i} = \alpha + \delta T_{i} + \gamma I_{i} + \beta'X_{i} + \epsilon_{i}$$
$$\Delta T_{i} = \omega + \tau_{1}E_{i} + \tau_{2}E_{i}I_{i} + \tau_{3}M_{i} + \tau_{4}M_{i}I_{i} + \mu I_{i} + \Gamma' X_{i} + \epsilon_{i} $$ 
where $T_{i}$ is enrollment dummy, $E_{i}$ is eligibility cutoff dummy, $I_{i}$ is index, $M_{i}$ is dummy for region where enrollment increases rapidly

@@

- RD results show that CCT increases deforestation
\vfill
- Why? What changes at the household level?
\vfill
- Approach 1: Use experimental sample (Progresa) with household survey data
\vfill
- Approach 2: Heterogeneity by road density to study role of market access

## Mechanisms: Consumption channel
- Before (1997-1998) and after (2000) data on consumption
- 506 villages, 320 treated, 186 control
- Treatment at the village level

@@

- $y_{it}$ = consumption
- cluster at locality level 
\vfill
**Hypothesis: $\gamma_{3}>0$**

@@

Heterogeneity by infrastructure quality:
\begin{align*} 
y_{it} & = \beta_{0} + \beta_{1}Treat_{i} + \beta_{2}Post_{t} + \beta_{3}Treat_{i} \times Post_{i} \\ & + \beta_{4}Road_{i} + \beta_{5} Road_{i} \times Treat_{i} + \beta_{6} Road_{i} \times Post_{t} \\ & + \beta_{7} Road_{i} \times Treat_{i} \times Post_{i} + \epsilon_{it}
\end{align*}

@@

```
\vfill
Bigger impact where supply response is localized

@@

## Thoughts?
- Form discussion group to think of alternate story

- Do we believe the story about increased demand for land-intensive goods?
  - Increased income relieves credit constraints
  - Expands capital, increases ag productivity on extensive margin b/c malfunctioning factor markets.
  - Market access may improve access to inputs, and reduce deforestation. 
  - Consistent with papers results.

- Only focused on one mechanism
  - returns to off-farm labor
  - income may increase aggregate production and lead to deforestation?
  - market access may lead to migration, less population pressure and deforestation
  - also consistent story

@@

- But where is supply response coming from?
  - Need much more details spatial analysis
  - SLX model

- Road density can be capturing other dimensions of underdevelopment, not only market access

- How do we know that number of plots, and farm size increases are encroaching into forest? what if they displace other land types

- Policy implication limited by thiessen polygon
  - What is the natural unit for policy implementation?

@@

@@

- Energy ladder: as income $\uparrow$, move from solid fuel $\rightarrow$ gas $\rightarrow$ electricity
  - Electricity may displace dirty energy (wood)

- But, electricity also increases total energy (fridge, AC)
  - energy footprint increases with income

@@

## GDP and Car Ownership
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.show='hold',  out.width = "45%", fig.align = "center"}

@@

## How can we explain the S-shape?

**Set up:**

@@

- Let $\bar{Y} = 1/2(Y_{1}+Y_{2})$ be average income

@@

- First period buyers respond only to first period income (lemma 1)

@@

- Prediction 0: Acquisition increasing in income
- Prediction 1: S-shaped curve in acquisition
- Prediction 2: Faster income growth leads to more period 2 adoption
- Prediction 3: Period 1 ownership depends on interaction of income and growth

## Empirical Setting: Oportunidades (again)

- 320 randomly selected communities given early treatment (April 1998)
\vfill
- 186 randomly selected communities given late treatment (October 1999)
\vfill
- This paper: N=506 communities, 10,000 households surveyed (1997-2007)
\vfill
- Compare early and late households
  - Similar b/c both groups selected on ``vulnerability'' characteristics
  - Strongly balanced on covariates

@@

$$h(a_{it}) = Pr(a_{it}=1 | a_{it-1}=0) = \gamma_{0} + \alpha_{1} CI_{it} + \beta X_{i} + \beta_{t}F_{i} + R_{rt} + \epsilon_{it}$$
where h(a) is the prob. that $i$ buys $a$ in time $t$ conditional on not having it in $t-1$. 
\vfill

- **Prediction 1:** S-shaped curve in acquisition
$$h(a_{it}) = Pr(a_{it}=1 | a_{it-1}=0) = \gamma_{0} + \alpha_{1} CI_{it} + \alpha_{2} CI^{2}_{it} + \beta X_{i} + R_{rt} + \epsilon_{it}$$
Hypothesis: $\alpha_{2} >0$

@@

- **Prediction 2 and 3:** conditional on having same level of cumulative income, households which accumulated income slower are less likely to acquire asset; this effect is increasing in cumulative income

$$h(a_{it}) = Pr(a_{it}=1 | a_{it-1}=0) = \gamma_{0} + \alpha_{1} CI_{it} + \alpha_{3} Early_{i} + \alpha_{4} (Early_{i} \times CI_{it}) + \beta X_{i} + R_{rt} + \epsilon_{it}$$
- where $Early_{i}=1$ if $i$ began receiving transfers 18 months before control households

@@

- Cumulative Income is sum of wage, farm, business income, CCT transfers
  - measurement error in wage data
  - cumulation requires interpolating between survey waves
\vfill
- Double counting problem if households invest CCT money into business
\vfill
- **Solution:** use cumulative CCT **transfers** to instrument cumulative **income**
  - Assumes transfers affect consumption only through income
  - Is this reasonable?

@@

@@

## Results: Prediction 2 and 3
- include only if time

@@

<!-- We learned about effect of income on energy. What about environment? -->

## Lessons from income/consumption channel?

- What do we learn about how development affects the environment?

- What are shortcomings of research designs? Measurement?

- External validity?

@@

- Does earned income have different effect than transfers?
\vfill
- Which externalities have highest income elasticities?
\vfill
- Can the effect of income on $MWTP_{e}$ be separated from changes in consumption?

@@

@@

- Development increases access to **productive** capital
  - How do we think about environmental implications?
\vfill
- Case I: Development increases agricultural productivity
  - Abman et al. (2023)
\vfill
- Case II: Development spurs infrastructure development
  - Asher et al. (2020)

@@

@@

- Continuum of agents with heterogeneous outside options
  - Changes in returns to ag affect number of agents who select into farming

@@

- Differentiating with respect to $\Omega$ gives:
$$\frac{\partial D^{*}}{\partial \Omega} = \underbrace{n^{*}\frac{\partial \gamma^{*}}{\partial \Omega}}_\text{$\leq 0$} + \underbrace{\gamma^{*}\frac{\partial n^{*}}{\partial \Omega}}_\text{$\geq 0$}$$

@@

- Agricultural survey on 7,781 households (451 villages) in 2011
  - Detailed cropping practices from last two seasons

@@

- Empirical issues
- Alternative explanations
- External validity (recall forest is private in Uganda)

@@

- Roads may facilitate agricultural labor loss
  - Will farmers turn to crop burning, despit the costs?

@@

## Data
- Roads: SHRUG database (N=11,151 villages w/o paved road at baseline)
\vfill
- Fires: NASA EODIS 
  -\# fires w/n 10km of village polygon
\vfill
- Pollution: Van donkelaar et al. (2016) measure PM2.5 at 10km resolution
\vfill
- Infant mortality: NFHS-IV (2015-16)
  - child-birthyear panel from 2001-2013
  - Identify survey clusters (villages) within 50km of each PMGSY village

@@

```
- Road access increases annual crop fires by 60\%
- Road access increase PM2.5 level by 0.5 $\mu g/m^{3}$ (1.1\%)

@@

## Mechanism: Roads induce labor exit
Back to RDD design:
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.show='hold',  out.width = "85%", fig.align = "center"}

@@

>>>>

---
title: "| Environmental and Development Economics\n| Module 2 - How does development affect the environment?
  \n"
author: |
  | Raahil Madhok
  | UMN Applied Economics
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: Montpellier
  slidy_presentation: default
  ioslides_presentation: default
classoption: aspectratio=169
header-includes:
- "\\AtBeginDocument{\\title[Module 2 - development and environment]{Environmental and Development
  Economics \\\\ Module 2 - How does development affect the environment?}}"
- \input{header.tex}
---

@@

- Replication project (Ryan announcement)
  - Due Oct 17th
\vfill
- Research proposal
  - Start thinking about idea + come talk to me
  - First draft: Oct 3rd (pass/fail)
\vfill
- Lecture slides will be posted before each class
\vfill
- Reading list is a work in progress
\vfill
- 5 min break mid-way through lectures

@@

- **Guiding question:** how does economic development affect the environment?
\vfill
- Descriptive overview
\vfill
- Channel I: Income effects (today)
  - **Changes in consumption**
  - **Energy**
  - Diet
\vfill
- Channel II: Technology and Infrastructure (next week)
  - Agricultural productivity
  - Infrastructure

\vfill
- Channel III: Institutions
  - Later in the course

@@

## GDP and Lead in Gasoline
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.show='hold',  out.width = "80%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/ekc_lead.png",
                        auto_pdf = TRUE
                        )

@@

- What variation in "development" can approximate the experiment?
  - what proxies development? (see \href{https://github.com/teamookla/ookla-open-data}{ookla speed test})
  - what does the proxy miss?
  - what (non-random) variation isolates causal relationship?

@@

- Poverty alleviation may $\uparrow$ demand for resource-intensive good
\vfill
- Or, it may $\uparrow$ demand for conservation
\vfill
- **This paper:** Exploit RDD eligibility to study impact of CCT on forests

@@

\vfill
- If we control for OVB, will panel fixed effects work? 
  <!-- - No: small year-to-year income changes do not reflect ``development'' -->
  <!-- - Predictable short-term income fluctuations different than permanent change -->

@@

- **Main variables:** village lat/lon, eligibility index, forest cover (satellite)
  - Enrollment by village until 2003
\vfill
- **Sample size:** 105,749 villages
\vfill
- **Unit of analysis:** only village centroids provided?
  - How do we estimate village-level impacts? Thiessen polygons
  - Area around a point where every location closer to the point than to all others
  - Problems?

@@

## Empirical Strategy: OLS Equation

$$\Delta f_{i} = \alpha + \delta T_{i} + \beta'X_{i} + \epsilon_{i}$$

  - where $\Delta f_{i}$ is \% deforestation in polygon i over 2000-2003
  - $T_{i}=1$ if the locality enrolled by 2003
  - $X_{i}$ includes poverty

- But $T_{i}$ is not random!

- If discontinuity is sharp, replace $T_{i}$ with $1[E_{i}>-1.2]$
- RDD: Sample window $-2 < I < -0.2$

## RDD Equation

- Discontinuity is fuzzy, not sharp
  - How do we know? Two reasons

- Solution: Use cutoff to instrument treatment probability

- Fuzzy RDD/IV:
$$\Delta f_{i} = \alpha + \delta T_{i} + \gamma I_{i} + \beta'X_{i} + \epsilon_{i}$$
$$\Delta T_{i} = \omega + \tau_{1}E_{i} + \tau_{2}E_{i}I_{i} + \tau_{3}M_{i} + \tau_{4}M_{i}I_{i} + \mu I_{i} + \Gamma' X_{i} + \epsilon_{i} $$ 
where $T_{i}$ is enrollment dummy, $E_{i}$ is eligibility cutoff dummy, $I_{i}$ is index, $M_{i}$ is dummy for region where enrollment increases rapidly

@@

- RD results show that CCT increases deforestation
\vfill
- Why? What changes at the household level?
\vfill
- Approach 1: Use experimental sample (Progresa) with household survey data
  - Progresa had randomized rollout at village level for first 3 yrs

\vfill
- Approach 2: Heterogeneity by road density to study role of market access

## Mechanisms: Consumption channel
- Before (1997-1998) and after (2000) data on consumption
- 506 villages, 320 treated, 186 control (not-yet-treated)
- Treatment at the village level

@@

- $y_{it}$ = consumption
- $Treat_{i}=1$ if household $i$ in treated locality
- cluster at locality level bc randomization was at locality level
- How do we interpret $\gamma_{3}$?
\vfill
**Hypothesis: $\gamma_{3}>0$**

@@

Heterogeneity by infrastructure quality:
\begin{align*} 
y_{it} & = \beta_{0} + \beta_{1}Treat_{i} + \beta_{2}Post_{t} + \beta_{3}(Treat_{i} \times Post_{t}) \\ & + \beta_{4}Road_{i} + \beta_{5} (Road_{i} \times Treat_{i}) + \beta_{6} (Road_{i} \times Post_{t}) \\ & + \beta_{7} (Road_{i} \times Treat_{i} \times Post_{t}) + \epsilon_{it}
\end{align*}

@@

```
\vfill
More deforestation in places with less market access (low road density)

@@

## Thoughts?

- Do we believe the story about increased demand for land-intensive goods?
  - Less market access means less access to inputs
  - Underinvest in technology, leading to higher deforestation
  - Maybe this is a story about technology, not consumption

- Can you think of an alternate explanation?

- Other concerns?

@@

@@

\centering
\huge Take a 5 minute break

@@

- Energy ladder: as income $\uparrow$, move from solid fuel $\rightarrow$ gas $\rightarrow$ electricity
  - Electricity may displace dirty energy (wood), which $\downarrow$ pollution

- But, electricity also increases total energy (fridge, AC)
  - ...which $\uparrow$ pollution
  - energy footprint increases with income

@@

## GDP and Car Ownership
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.show='hold',  out.width = "40%", fig.align = "center"}

@@

```
Of course, these are correlations

## But first, some theory

- Imagine we can exogenously vary income to poor households
\vfill
- Will transfers to initially poor households yield **lumpy** investments?
  - Movements up the ladder involve big purchases
  - May explain why 1.3 billion people lack electricity (Gertler et. al, 2016) 
\vfill
- Credit constraints play an important role
  - Without credit markets, buy now but have nothing in next period
  - With savings, can save now, buy later (smooth consumption)
  - With borrowing, buy now, smooth consumption over time
\vfill
- Lets formalize this

## Model Set up

@@

- Define $\bar{Y} = 1/2(Y_{1}+Y_{2})$ = average per period income

@@

- Higher income in any period leads savers to buy more

- But, first period buyers respond only to first period income (lemma 1)

@@

- Prediction 0: Acquisition increasing in income
- Prediction 1: S-shaped curve in acquisition
<!-- - Prediction 2: Faster income growth leads to more period 2 adoption -->

## Empirical Setting: Mexico Oportunidades (again)

- 320 randomly selected communities given early treatment (April 1998)
  - 52\% were ``eligible'' for cash transfer
\vfill
- 186 randomly selected communities given late treatment (October 1999)
\vfill
- This paper: N=506 communities, 10,000 households surveyed (1997-2007)
  - Surveys done in  1998, 1999, 2000, 2003, and 2007
\vfill
- Compare early and late treatment households (**not treatment vs. control**)
  - Similar b/c both groups selected on ``vulnerability'' characteristics
  - Strongly balanced on covariates
  - Assumes timing is random

@@

$$h(a_{it}) = Pr(a_{it}=1 | a_{it-1}=0) = \gamma_{0} + \alpha_{1} CI_{it} + \beta X_{i} + \beta_{t}F_{i} + R_{rt} + \epsilon_{it}$$

  - where h(a) is the prob. that $i$ buys $a$ in time $t$ conditional on not having it in $t-1$. 
  - $CI_{it}$ is cumulative income
  - $X_{i}$ includes baseline household controls
  - $F_{i}$ are family structure variables interacted w/ survey round
  - $R_{rt}$ are state-year fixed effects

Hypothesis: $\alpha_{1} >0$

@@

- **Prediction 1:** S-shaped curve in acquisition
$$h(a_{it}) = Pr(a_{it}=1 | a_{it-1}=0) = \gamma_{0} + \alpha_{1} CI_{it} + \alpha_{2} CI^{2}_{it} + \beta X_{i} + R_{rt} + \epsilon_{it}$$

Hypothesis: $\alpha_{2} \lessgtr 0?$

<!-- ## Empirical Design -->

<!-- - **Prediction 2:** conditional on having same level of cumulative income, households which accumulated income slower are less likely to acquire asset -->

<!-- $$h(a_{it}) = Pr(a_{it}=1 | a_{it-1}=0) = \gamma_{0} + \alpha_{1} CI_{it} + \alpha_{3} Early_{i} + \alpha_{4} (Early_{i} \times CI_{it}) + \beta X_{i} + R_{rt} + \epsilon_{it}$$ -->

<!-- - where $Early_{i}=1$ if $i$ began receiving transfers 18 months before control households -->
<!-- - $Early_{i}$ is synonymous with treated, and $Late_{i}$ is control (not-yet-treated) -->

@@

- Cumulative Income is sum of wage, farm, business income, CCT transfers
  - Reporting biases in wage data
  - Income reported in survey year, not cumulative
  - Authors interpolate across years, adding noise
\vfill
- Double counting problem if households invest CCT money into business
\vfill
- **Solution:** use potential CCT **transfers** to instrument cumulative **income**

@@

- What is the exclusion restriction?
- Is it reasonable?

@@

@@

@@

- Does earned income have different effect than transfers?
\vfill
- Can we trace out the full S-curve?
\vfill
- Is there an S-curve for other technologies?
\vfill
- Which externalities have highest income elasticities?
\vfill

@@

- Research proposal first draft: Oct 3rd
  - If you find idea w/n 10 days, gives 2 weeks to write
\vfill
- Class presentations: separate Canvas upload each week
\vfill
- Replication project?

## Recap: Alix-Garcia et al. (2013)

- Do we believe the story about increased demand for land-intensive goods?
  - Less market access means less access to inputs
  - Underinvest in technology, leading to higher deforestation
  - Maybe this is a story about technology, not consumption

- Can you think of an alternate explanation?
  - Talk to your neighbor for five minutes

- Other concerns?

@@

- Development increases access to **productive** capital
  - How do we think about environmental implications?
\vfill
- Case I: Development increases agricultural productivity
  - Abman et al. (2023)
\vfill
- Case II: Development spurs infrastructure development
  - Garg et al. (2023)

@@

## Existing Work

- Theory: (Angelsen and Kaimowitz,2001; Balsdon, 2007; Goldstein et al., 2012; Green et al., 2005; Phalan et al.,2016; Takasaki,2006). 
  - Most models yield ambiguity
\vfill
- Most work on agricultural productivity and deforestation
  - Intensive margin dominates: Abman et al. (2024), Abman and Carney (2020a), Abman and Carney (2020b), Assuncao et al. (2023)
\vfill
- New work: market access and deforestation 
  - Abman and Lundberg (2024)
\vfill
- Open Topics:
  - Effect of agriculture on deforestation in the first place? (Madhok et al. 2024)
  - Role of factor markets? Labor vs. input markets
  - Mechanisms behind market access effects
  - Agriculture, migration and deforestation

@@

- Continuum of agents with heterogeneous outside options
  <!-- - Changes in returns to ag affect number of agents who select into farming -->

@@

- Differentiating with respect to $\Omega$ gives:
$$\frac{\partial D^{*}}{\partial \Omega} = \underbrace{n^{*}\frac{\partial \gamma^{*}}{\partial \Omega}}_\text{$\lessgtr 0$} + \underbrace{\gamma^{*}\frac{\partial n^{*}}{\partial \Omega}}_\text{$\lessgtr 0$}$$

@@

- BRAC agricultural survey on 7,781 households (451 villages) in 2011
  - Detailed cropping practices from last two seasons

@@

- Empirical issues
- Alternative explanations
- External validity

@@

- Roads may facilitate agricultural labor loss
  - Will farmers turn to crop burning, despite the costs?

@@

## Data
- Roads: SHRUG database 
  - N=11,151 villages w/o paved road at baseline (2001)
  - Main variable: road construction date in village
\vfill
- Fires: NASA EODIS
  - \# fires w/n 10km of village polygon
\vfill
- Pollution: Van donkelaar et al. (2016) measure PM2.5 at 10km resolution
  - Black Carbon and Organic Carbon from biomass/non-biomass sources
\vfill
- Infant mortality: NFHS-IV (2015-16)
  - child-birthyear panel from 2001-2013
  - Identify survey clusters (villages) within 50km of each PMGSY village

@@

```
- Note: optimal bandwidth = +/- 84 around population threshold
- Road access increases annual crop fires by 60\%
- Road access increase PM2.5 level by 0.5 $\mu g/m^{3}$ (1.1\%)

@@

## Mechanism: Roads induce labor exit
- Back to RDD design
- Data from Socioeconomic and Caste Census (2013), provided in SHRUG
- Outcome: ag. labor share in districts with high/low ag wage
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.show='hold',  out.width = "85%", fig.align = "center"}

@@

## Next Time
